---
title: "R Functions Lab (Class 06)"
author: "Eddie Cintora (A17862193)"
date: "April 16, 2026"
format: pdf
---

### Q1. Write your first R function: add()

```{r}
## Q1. Write your first R function: add()

### Q1a
add <- function(x, y) {
  x + y
}

add(4, 7)
```

Q1b

```{r}
add <- function(x, y = NULL) {
  if (is.null(y)) {
    sum(x)
  } else {
    x + y
  }
}

add(4, 7)
add(c(4, 7, 10))
```

Q1c

```{r}
add <- function(...) {
  sum(...)
}

add(4, 7)
add(c(4, 7, 10))
add(1, 2, 3, -4)
```

### Q2. Write a generate_dna() function

```{r}
generate_dna <- function(n) {
  nts <- c("A", "T", "G", "C")
  sample(nts, size = n, replace = TRUE)
}

generate_dna(6)
```

Q2b

```{r}
generate_dna <- function(n, as_string = TRUE) {
  nts <- c("A", "T", "G", "C")
  seq_vec <- sample(nts, size = n, replace = TRUE)
  
  if (as_string) {
    paste(seq_vec, collapse = "")
  } else {
    seq_vec
  }
}

generate_dna(6, as_string = FALSE)
generate_dna(11)
```

Q2c

```{r}
generate_dna <- function(n) {
  nts <- c("A", "T", "G", "C")
  seq_vec <- sample(nts, size = n, replace = TRUE)
  seq_string <- paste(seq_vec, collapse = "")
  
  cat(paste0(">len", n, "\n", seq_string, "\n"))
}

generate_dna(11)
```

### Q3. Write a generate_protein() function

```{r}
generate_protein <- function(n) {
  aa <- c("A", "R", "N", "D", "C", "E", "Q", "G", "H", "I",
          "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V")
  
  seq_vec <- sample(aa, size = n, replace = TRUE)
  paste(seq_vec, collapse = "")
}

generate_protein(6)
```

### Q4. Generate random protein sequences of length 6 to 13

```{r}
for (i in 6:13) {
  cat(paste0(">id.", i, "\n"))
  cat(generate_protein(i), "\n")
}
```

### Q5. BLASTp search against nr — are your peptides “unique in nature”?

**Sequence generated utilized for BLAST results:**

\>id.6\
TDAWPL\
\>id.7\
YGLNYKY\
\>id.8\
ARNKIRME\
\>id.9\
IHRNVMDTW\
\>id.10\
GCLIHIQYVT\
\>id.11\
MLACGVSTCHI\
\>id.12\
RCCESASQEEKI\
\>id.13\
NEGICNLWGQSTH

**Results Table**

| Length (aa) | Best hit % identity | Best hit % coverage | Unique? (Y/N) |
|-------------|---------------------|---------------------|---------------|
| 6           | 100%                | 100%                | N             |
| 7           | 100%                | 100%                | N             |
| 8           | 100%                | 100%                | N             |
| 9           | 87.5%               | 89%                 | Y             |
| 10          | 90.0%               | 90%                 | Y             |
| 11          | 90.0%               | 91%                 | Y             |
| 12          | 66.7%               | 100%                | Y             |
| 13          | 88.9%               | 69%                 | Y             |

Q5a: Random peptides will appear “unique in nature” when approachinf the approximate 9 amino acid length. Shorter sequences have displayed a trend in perfect matches, while longer sequences usually did not.

Q5b: Peptides very short in length are frequently found because the total sequence space utilized is small. The number of possible sequences is 20\^L, so as length begins to increase, the number of possible sequences grows exponentially with it. This begins to exceed the number of known proteins very fast, making exact matches unlikely.

### Q6. Connecting your findings to immunology (MHC class II and T-cell activation)

The minimum peptide length required for MHC class II presentation is most likely to be around 9–12 amino acids. Shorter peptides are not a great choice because they are too common in length and may appear in both self and non-self proteins. This would make it hard for the immune system to distinguish between them. Longer peptides occupy a much larger sequence space, making them more unique and better suited for identifying foreign antigens. This improves the specificity of immune recognition and helps ensure appropriate T-cell activation.
